\documentclass[11pt,a4paper]{article}
\usepackage[utf8]{inputenc}
\usepackage{geometry}
\geometry{margin=1in}
\usepackage{hyperref}
\usepackage{enumitem}
\usepackage{listings}
\usepackage{xcolor}

\lstset{
  language=C++,
  basicstyle=\ttfamily\small,
  keywordstyle=\bfseries,
  commentstyle=\itshape,
  showstringspaces=false,
  breaklines=true,
  captionpos=b,
  numbers=left,
  numberstyle=\tiny\color{gray},
}

\title{Digital Design Training Program \\ \large Day 4: Timekeeping Logic \& User Interface}
\author{Authored by }
\date{April 1, 2026}

\begin{document}

\maketitle

\section*{Theme: Implementing the core clock logic, handling user inputs robustly, and building the Edit Mode state machine.}

\subsection*{Module 4.1: Cascading Time Logic (45 Minutes)}

\begin{itemize}
    \item \textbf{Goal:} Convert the raw 10Hz background ``tick'' from the Timer1 ISR into a human-readable 24-hour clock format and display it.
    \item \textbf{Conceptual Overview:} The ISR gives us a steady pulse, incrementing \verb|deciseconds|. In our main loop, we will constantly check this variable. When it reaches 10, we know a full second has passed. This triggers a ``cascade'' where we reset the \verb|deciseconds| and increment \verb|seconds|. The same logic then applies to turn seconds into minutes, minutes into hours, and hours into days (by rolling over at 24).
    \item \textbf{Action Steps \& Code:}
\begin{lstlisting}[caption={Code for Module 4.1: Full Clock Logic}]
// We need all the previous parts: BCD, digits, and the timer ISR
const int bcd_pins[] = {8, 9, 10, 11};
const int digit_pins[] = {A0, A1, A2, A3, A4, A5};
const int N_DIGITS = 6;

// --- Time Variables ---
volatile int deciseconds = 0; // Updated by ISR
int seconds = 0;
int minutes = 0;
int hours = 12; // Start at 12:00:00

// This array is the ``single source of truth'' for the display.
int time_arr[N_DIGITS] = {1, 2, 0, 0, 0, 0};

// This function takes the time variables and maps them to the display array.
void map_time_to_display_array() {
  time_arr[0] = hours / 10;   // Tens-of-hours
  time_arr[1] = hours % 10;   // Single hours
  time_arr[2] = minutes / 10; // Tens-of-minutes
  time_arr[3] = minutes % 10; // Single minutes
  time_arr[4] = seconds / 10; // Tens-of-seconds
  time_arr[5] = seconds % 10; // Single seconds
}

// This function contains the core cascading logic.
void update_time() {
  // Check if a second has passed.
  if (deciseconds >= 10) {
    deciseconds = 0; // Reset for the next second
    seconds++;
    if (seconds >= 60) {
      seconds = 0;
      minutes++;
      if (minutes >= 60) {
        minutes = 0;
        hours++;
        if (hours >= 24) {
          hours = 0;
        }
      }
    }
    // After any change, we must update the display array.
    map_time_to_display_array();
  }
}

// ISR remains the same, just increments deciseconds
ISR(TIMER1_COMPA_vect) {
  deciseconds++;
}

void loop() {
  // In the main loop, we constantly check and update the time,
  // and then refresh the display.
  update_time();
  mux_update(); // The multiplexing function from Day 3
}
\end{lstlisting}
    \item \textbf{Checkpoint:} Your clock is now a fully functional 24-hour clock. The display is stable and the time updates accurately every second, driven by the hardware interrupt.
\end{itemize}
\newpage
\subsection*{Module 4.2: Control Buttons \& Non-Blocking Debounce (45 Minutes)}

\begin{itemize}
    \item \textbf{Goal:} Read physical user inputs reliably without blocking the processor or registering false ``double-clicks''.
    \item \textbf{Requirements:} 4x Pushbuttons (e.g., Pause/Play, Select, Increment, Decrement).
    \item \textbf{Conceptual Overview:} Mechanical buttons are messy. When you press one, the metal contacts ``bounce'' for a few milliseconds, creating a rapid series of on-off signals. A naive \verb|digitalRead()| would see this as multiple presses. We need a non-blocking way to (1) detect only the first press and (2) ignore the subsequent bounces.
    \item \textbf{The Debounce Strategy:}
    \begin{enumerate}
        \item \textbf{State Change Detection:} Don't act just because a button is \verb|LOW|. Act only when it \emph{changes} from \verb|HIGH| to \verb|LOW|. This requires storing the button's \verb|last_state|.
        \item \textbf{Timing:} After detecting a valid \verb|HIGH| to \verb|LOW| transition, record the time using \verb|millis()|. Ignore any further transitions on that button until a certain amount of time (e.g., 50ms) has passed.
    \end{enumerate}
    \item \textbf{Action Steps \& Code:}
\begin{lstlisting}[caption={Code for Module 4.2: A Robust Button Handler}]
const int BTN_PIN = 4; // Example for one button

int last_btn_state = HIGH; // Assume button is not pressed initially
unsigned long last_debounce_time = 0;
const unsigned long DEBOUNCE_DELAY = 50; // 50ms

void setup() {
  Serial.begin(9600);
  pinMode(BTN_PIN, INPUT_PULLUP);
}

void check_button() {
  int current_reading = digitalRead(BTN_PIN);

  // Check for a state change: was HIGH, is now LOW.
  if (current_reading == LOW && last_btn_state == HIGH) {
    // Check if enough time has passed since the last valid press.
    if (millis() - last_debounce_time > DEBOUNCE_DELAY) {
      // --- THIS IS WHERE YOU PUT YOUR ACTION ---
      Serial.println(``Button Pressed!'');
      // --- END OF ACTION ---

      // Record the time of this valid press.
      last_debounce_time = millis();
    }
  }
  
  // Save the current state for the next loop iteration.
  last_btn_state = current_reading;
}

void loop() {
  // In the main loop, you just call your button checker.
  // It doesn't block the rest of your code.
  check_button();
  // Your multiplexing code would also go here.
}
\end{lstlisting}
    \item \textbf{Checkpoint:} Open the Serial Monitor. Press and hold the button. The message ``Button Pressed!'' appears only once. Release and press again quickly. It still only registers one press per distinct push.
\end{itemize}
\newpage
\subsection*{Module 4.3: The Edit Mode State Machine (60 Minutes)}

\begin{itemize}
    \item \textbf{Goal:} Architect a robust software state machine that allows the user to pause the clock, select a digit, and modify the time safely.
    \item \textbf{Conceptual Overview:} We will define two main states for our clock: \verb|RUNNING| and \verb|EDITING|. The program will behave differently depending on which state it's in. This is the foundation of almost all interactive software.
    \item \textbf{Action Steps \& Code:}
    \begin{enumerate}
        \item \textbf{State Definition:} Use an \verb|enum| to create a new data type for our states. This makes the code much more readable than using numbers.
        \begin{lstlisting}
enum Mode { RUNNING, EDITING };
Mode current_mode = RUNNING; // Start in RUNNING mode
int cursor_position = 0; // Which digit is selected (0-5)
        \end{lstlisting}
        \item \textbf{Mode Toggling:} The Pause/Play button will flip the \verb|current_mode|. When entering \verb|EDITING| mode, we stop the ISR from incrementing time.
        \item \textbf{Visual Feedback (Blinking):} In \verb|EDITING| mode, we need a blinking cursor. We do this with another non-blocking \verb|millis()| check inside the multiplexing loop. If the cursor is on the current digit, we simply skip turning it on for half a second, making it disappear.
        \item \textbf{Data Integrity (Bounds Checking):} This is critical. When the user presses ``Increment'' on the tens-of-minutes digit, it must roll over from 5 back to 0, not go to 6. This requires a \verb|switch| statement or \verb|if/else| chain based on the \verb|cursor_position|.
    \end{enumerate}
\begin{lstlisting}[caption={Code for Module 4.3: Simplified Edit Logic}]
// Inside your button handler, for the Increment button:
if (current_mode == EDITING) {
  // Get the current value of the two digits being edited
  int* tens_digit = &time_arr[cursor_position - (cursor_position % 2)];
  int* ones_digit = &time_arr[cursor_position - (cursor_position % 2) + 1];
  
  // Example for minutes
  if (cursor_position == 2 || cursor_position == 3) {
      int current_mins = (*tens_digit * 10) + *ones_digit;
      current_mins = (current_mins + 1) % 60; // Increment and wrap around 60
      *tens_digit = current_mins / 10;
      *ones_digit = current_mins % 10;
  }
  // Add similar logic for hours (0-23) and seconds (0-59)
}
\end{lstlisting}
    \item \textbf{The Main Loop Re-architected:}
\begin{lstlisting}
// The ISR is modified to check the mode
ISR(TIMER1_COMPA_vect) {
  if (current_mode == RUNNING) {
    deciseconds++;
  }
}

void loop() {
  // 1. Handle Inputs
  check_all_buttons(); // This function might change our mode or cursor

  // 2. Update State
  if (current_mode == RUNNING) {
    update_time(); // The cascading logic from 4.1
  }
  
  // 3. Update Display
  mux_update(); // The multiplexer now contains the blinking logic
}
\end{lstlisting}
    \item \textbf{Checkpoint:} The clock is now fully interactive. You can pause it, the selected digit blinks, you can move the cursor, and incrementing/decrementing the time follows real-world rules (e.g., minutes don't go past 59). You can then resume the clock, and it continues from the new time you set.
\end{itemize}

\end{document}
