\documentclass[11pt,a4paper]{article}
\usepackage{xcolor}
\usepackage[utf8]{inputenc}
\usepackage{geometry}
\geometry{margin=1in}
\usepackage{hyperref}
\usepackage{enumitem}
\usepackage{amsmath}
\usepackage{listings}

\lstset{
  language=C,
  basicstyle=\ttfamily\small,
  keywordstyle=\bfseries,
  commentstyle=\itshape,
  showstringspaces=false,
  breaklines=true,
  captionpos=b,
  numbers=left,
  numberstyle=\tiny\color{gray},
}

\title{Digital Design Training Program \\ \large Day 10: Capstone Integration and Optimization}
\author{Authored by }
\date{April 1, 2026}

\begin{document}

\maketitle

\section*{Theme: Fusing the hardware drivers, the RK4 math engine, and the parser into a cohesive, physical scientific calculator.}

\subsection*{Module 10.1: The Hardware/Software Pipeline (90 Minutes)}

\begin{itemize}
    \item \textbf{Goal:} Link the keypad driver, LCD driver, and parser to create a reactive user interface that captures user input and triggers the calculation pipeline.
    \item \textbf{Conceptual Overview:} The system will be interrupt-driven. The main loop will be idle. The Keypad ISR will be the star, capturing characters, echoing them to the LCD, and setting a flag when the `=' key is pressed. This flag signals the main loop to wake up and perform the calculation.
    \item \textbf{Action Steps \& Code:}
    \begin{enumerate}
        \item \textbf{Keypad Mapping:} We define a 2D array that maps the physical row/column of the keypad to a character. This makes the ISR code clean and easy to modify.
\begin{lstlisting}
const char KEYMAP[5][5] = {
    {`s',`c',`t',`^',`/'}, // sin, cos, tan, pow, div
    {`7',`8',`9',`*',`('},
    {`4',`5',`6',`-',`)'},
    {`1',`2',`3',`+',`l'}, // ln
    {`0',`.',`e',`=',`C'}  // exp, equals, clear
};
\end{lstlisting}
        \item \textbf{The Global Buffers \& Flags:} We need a few global \verb|volatile| variables to communicate between the ISR and the main loop.
\begin{lstlisting}
#define BUF_SIZE 32
volatile char g_input_buffer[BUF_SIZE];
volatile int g_buffer_index = 0;
volatile uint8_t g_start_evaluation = 0; // Flag
\end{lstlisting}
        \item \textbf{The Keypad ISR (Upgraded):} The ISR from Day 6 is upgraded. After finding the \verb|key|, it doesn't just store it. It acts as a UI controller.
\begin{lstlisting}
ISR(PCINT2_vect) {
    // ... (keypad scanning logic to find `key') ...
    
    if (key) { // If a valid key was found
        if (key == `=') {
            g_start_evaluation = 1; // Set the flag for the main loop
        } else if (key == `C') {
            g_buffer_index = 0;
            g_input_buffer[0] = `\0';
            lcd_cmd(0x01); // Clear screen command
        } else if (g_buffer_index < BUF_SIZE - 1) {
            // Add char to buffer and echo to LCD
            g_input_buffer[g_buffer_index++] = key;
            g_input_buffer[g_buffer_index] = `\0';
            lcd_data(key); // Print the character
        }
    }
    
    // ... (re-enable interrupts) ...
}
\end{lstlisting}
        \item \textbf{The Main Loop:} The main loop is now just a simple state checker.
\begin{lstlisting}
int main(void) {
    // ... (all initializations: lcd, keypad, interrupts) ...
    sei(); // Enable global interrupts

    while (1) {
        // Wait for the ISR to tell us to start
        if (g_start_evaluation) {
            cli(); // Disable interrupts during calculation
            g_start_evaluation = 0; // Reset the flag
            
            // --- TRIGGER THE PIPELINE ---
            // 1. Tokenize g_input_buffer
            // 2. Shunt the tokens to RPN
            // 3. Evaluate the RPN
            // 4. Display the result
            // --------------------------
            
            sei(); // Re-enable interrupts
        }
    }
}
\end{lstlisting}
    \end{enumerate}
    \item \textbf{Checkpoint:} You can type an equation like \verb|12+s(3)| on the keypad, see it appear on the LCD, and pressing \verb|=| makes the main loop wake up to begin processing.
\end{itemize}
\newpage
\subsection*{Module 10.2: RPN Evaluation \& The Math Engine (60 Minutes)}

\begin{itemize}
    \item \textbf{Goal:} Write the final piece of the puzzle: a function that evaluates the RPN queue produced by the Shunting-Yard algorithm.
    \item \textbf{Conceptual Overview:} Evaluating RPN is the simplest part. We use one stack (a \verb|FloatStack|). We read the RPN queue from left to right. If we see a number, we push it onto our stack. If we see an operator or function, we pop the required number of operands from the stack, perform the math, and push the single result back onto the stack. When the queue is empty, the single number remaining on the stack is our final answer.
    \item \textbf{Action Steps \& Code:}
\begin{lstlisting}[caption={Code for Module 10.2: The RPN Evaluator}]
// --- Assume all drivers, stacks, and rk4_solve are included ---

float evaluate_rpn(Token* rpn_queue, int queue_len) {
    FloatStack eval_stack;
    fstack_init(&eval_stack);

    for (int i = 0; i < queue_len; i++) {
        Token t = rpn_queue[i];

        if (t.type == T_NUMBER) {
            // If it's a number, convert from string to float and push
            fstack_push(&eval_stack, atof(t.value));
        } else if (t.type == T_OPERATOR) {
            // Operator: pop two operands, calculate, push result
            float b = fstack_pop(&eval_stack);
            float a = fstack_pop(&eval_stack);
            switch (t.value[0]) {
                case `+': fstack_push(&eval_stack, a + b); break;
                case `-': fstack_push(&eval_stack, a - b); break;
                case `*': fstack_push(&eval_stack, a * b); break;
                case `/': fstack_push(&eval_stack, a / b); break;
                // Power (a^b) is tricky. We can use logs: a^b = exp(b * ln(a))
                case `^': fstack_push(&eval_stack, rk4_solve(0, 1, b * rk4_solve(1, 0, a, deriv_ln), deriv_exp)); break;
            }
        } else if (t.type == T_FUNCTION) {
            // Function: pop one operand, call math engine, push result
            float a = fstack_pop(&eval_stack);
            if (strcmp(t.value, ``exp'') == 0) {
                fstack_push(&eval_stack, rk4_solve(0, 1, a, deriv_exp));
            } else if (strcmp(t.value, ``ln'') == 0) {
                fstack_push(&eval_stack, rk4_solve(1, 0, a, deriv_ln));
            }
            // Add sin, cos, etc. using their own derivative functions
        }
    }
    // The final result is the only thing left on the stack
    return fstack_pop(&eval_stack);
}
\end{lstlisting}
    \item \textbf{Checkpoint:} Your calculator is complete. Typing \verb|1+2*3=| correctly tokenizes, shunts to \verb|1 2 3 * +|, evaluates to \verb|7.0|, and displays it on the screen.
\end{itemize}
\newpage
\subsection*{Module 10.3: UI State Machines \& Benchmarking (60 Minutes)}

\begin{itemize}
    \item \textbf{Goal:} Add professional features like a SHIFT key and analyze the performance and accuracy of the finished product.
    \item \textbf{Action Steps:}
    \begin{enumerate}
        \item \textbf{Implementing a SHIFT key:}
        \begin{itemize}
            \item Add a new global state: \verb|volatile uint8_t g_shift_mode = 0;|
            \item Create a second keymap: \verb|const char KEYMAP_SHIFT[5][5] = ...| with \verb|asin|, \verb|acos|, \verb|sqrt|, etc.
            \item In the ISR, if the pressed key is \verb|SHIFT|, simply toggle \verb|g_shift_mode| and update the LCD with a ``SHIFT'' indicator.
            \item When looking up the key, use an if-statement:\\ \verb|if (g_shift_mode) { key = KEYMAP_SHIFT[r][c]; } else { key = KEYMAP[r][c]; }|.
        \end{itemize}
        \item \textbf{Benchmarking Speed:} To see how fast your code is, use a hardware timer. Before calling the calculation pipeline, reset a global \verb|volatile| millisecond counter to 0 and enable the timer. After the result is calculated, disable the timer. The value in the counter is your execution time.
        \item \textbf{Analyzing Accuracy:} Compare your calculator's results for complex equations (e.g., \verb|sin(0.5)^2 + cos(0.5)^2|) against a known good calculator. Where do they differ? This is due to floating point precision and the approximation error from the RK4 method.
        \item \textbf{Tuning the Engine:} In your \verb|rk4_solve| function, change the step size \verb|h|.
        \begin{itemize}
            \item A smaller \verb|h| (e.g., \verb|0.001|) will yield more accurate results but will take much longer to compute.
            \item A larger \verb|h| (e.g., \verb|0.1|) will be very fast but less accurate.
            \item This is the fundamental trade-off in numerical computing: \textbf{Accuracy vs. Speed}.
        \end{itemize}
    \end{enumerate}
    \item \textbf{Final Checkpoint:} You have successfully designed, built, and tested a complete scientific calculator from the ground up. You understand the entire pipeline, from the physics of the button contacts to the hardware drivers, the memory management, the parsing algorithms, the numerical approximation methods, and the final display.
\end{itemize}

\end{document}
