\documentclass[11pt,a4paper]{article}
\usepackage{xcolor}
\usepackage[utf8]{inputenc}
\usepackage{geometry}
\geometry{margin=1in}
\usepackage{hyperref}
\usepackage{enumitem}
\usepackage{amsmath}
\usepackage{listings}

\lstset{
  language=C,
  basicstyle=\ttfamily\small,
  keywordstyle=\bfseries,
  commentstyle=\itshape,
  showstringspaces=false,
  breaklines=true,
  captionpos=b,
  numbers=left,
  numberstyle=\tiny\color{gray},
}

\title{Digital Design Training Program \\ \large Day 7: Custom LCD Drivers and 8-Bit Architecture Limits}
\author{Authored by }
\date{April 1, 2026}

\begin{document}

\maketitle

\section*{Theme: Building a bare-metal communication driver for external displays and understanding the memory/processing constraints of an 8-bit microcontroller.}

\subsection*{Module 7.1: The 4-Bit Bare-Metal LCD Driver (120 Minutes)}

\begin{itemize}
    \item \textbf{Goal:} Write a complete C driver from scratch to command a 16x2 HD44780 LCD, mastering the concept of bit-shifting data into ``nibbles'' to save microcontroller pins.
    \item \textbf{Conceptual Overview:} The HD44780 is a classic character LCD. We could use 8 data lines to communicate with it, but we can save 4 precious I/O pins by using 4-bit mode. This means we have to send each 8-bit command or character as two separate 4-bit chunks (a high nibble and a low nibble).
    \item \textbf{Hardware Setup:} Wire the LCD to \verb|PORTB| (Digital Pins 8-13) to keep \verb|PORTC| and \verb|PORTD| free for the keypad.
    \begin{itemize}
        \item \textbf{RS (Register Select):} \verb|PB0| (Pin 8). HIGH for data, LOW for commands.
        \item \textbf{EN (Enable):} \verb|PB1| (Pin 9). We will pulse this to tell the LCD to read the data lines.
        \item \textbf{D4 to D7 (Data Lines):} \verb|PB2| to \verb|PB5| (Pins 10 to 13).
        \item \textbf{R/W:} Tie to Ground. We will only write to the LCD.
    \end{itemize}
    \item \textbf{Action Steps \& Code:}
\begin{lstlisting}[caption={Code for Module 7.1: Bare-Metal LCD Driver}]
#include <avr/io.h>
#include <util/delay.h>

// Pin definitions for clarity
#define LCD_PORT PORTB
#define LCD_DDR  DDRB
#define LCD_RS   PB0
#define LCD_EN   PB1
#define LCD_D4   PB2
#define LCD_D5   PB3
#define LCD_D6   PB4
#define LCD_D7   PB5

// Create masks for easy manipulation of control and data pins
#define LCD_DATA_PINS ((1<<LCD_D4)|(1<<LCD_D5)|(1<<LCD_D6)|(1<<LCD_D7))
#define LCD_CTRL_PINS ((1<<LCD_RS)|(1<<LCD_EN))

// Pulse the EN pin to latch data into the LCD
void lcd_pulse_en(void) {
    LCD_PORT |= (1 << LCD_EN);  // EN HIGH
    _delay_us(1);               // Wait >450ns
    LCD_PORT &= ~(1 << LCD_EN); // EN LOW
    _delay_us(50);              // Wait for command to complete
}

// Send a 4-bit nibble to the data pins
void lcd_write_nibble(uint8_t nibble) {
    // Clear the data pins first
    LCD_PORT &= ~LCD_DATA_PINS;
    // This is a more efficient way to set the pins than 4 if-statements.
    // It maps the high 4 bits of \verb|nibble| to the 4 data pins.
    LCD_PORT |= ((nibble & 0xF0) >> (4 - LCD_D4));
    lcd_pulse_en();
}

// Send a full 8-bit byte (command or data)
void lcd_write(uint8_t byte, uint8_t is_data) {
    if (is_data) {
        LCD_PORT |= (1 << LCD_RS); // RS HIGH for data
    } else {
        LCD_PORT &= ~(1 << LCD_RS); // RS LOW for commands
    }
    
    lcd_write_nibble(byte & 0xF0);      // Send high nibble
    lcd_write_nibble((byte << 4) & 0xF0); // Send low nibble
}

// --- Public API ---
void lcd_cmd(uint8_t cmd) { lcd_write(cmd, 0); }
void lcd_data(uint8_t data) { lcd_write(data, 1); }
void lcd_print(const char *str) { while (*str) lcd_data(*str++); }

void lcd_init(void) {
    // Set all control and data pins as outputs
    LCD_DDR |= (LCD_DATA_PINS | LCD_CTRL_PINS);
    _delay_ms(50); // Wait for LCD to power up

    // --- The Magic Initialization Sequence (from datasheet) ---
    lcd_write_nibble(0x30); _delay_ms(5);   // Wake up 1
    lcd_write_nibble(0x30); _delay_us(150); // Wake up 2
    lcd_write_nibble(0x30); _delay_us(150); // Wake up 3
    lcd_write_nibble(0x20); _delay_us(150); // Set to 4-bit mode

    lcd_cmd(0x28); // Function Set: 4-bit, 2-line, 5x8 font
    lcd_cmd(0x0C); // Display Control: Display ON, Cursor OFF
    lcd_cmd(0x01); // Clear display
    _delay_ms(2);  // Clear command takes longer
    lcd_cmd(0x06); // Entry Mode: Increment cursor, no shift
}

int main(void) {
    lcd_init();
    lcd_print(``Hello, Bare-Metal!'');
    while(1) {}
    return 0;
}
\end{lstlisting}
    \item \textbf{Checkpoint:} The screen successfully displays ``Hello, Bare-Metal!''. You have written a complete communication driver from scratch.
\end{itemize}
\newpage
\subsection*{Module 7.2: 8-Bit Math Limits \& PROGMEM (30 Minutes)}

\begin{itemize}
    \item \textbf{Goal:} Understand the high cost of floating-point math on an 8-bit AVR and learn to conserve RAM by storing constant data in Flash memory.
    \item \textbf{Conceptual Overview:} The ATmega328P is a simple 8-bit integer CPU. It has no hardware Floating Point Unit (FPU). When you use a \verb|float|, the \verb|avr-gcc| compiler must include large software libraries to simulate decimal math. This is slow and consumes a lot of program space. Likewise, the 2KB of SRAM is tiny; storing large, constant lookup tables or strings in SRAM is wasteful.
    \item \textbf{Action Steps:}
    \begin{enumerate}
        \item \textbf{The Cost of Floats:} Compile a simple program that does nothing. Note the size. Now add one line: \verb|volatile float x = 1.23 * 4.56;|. Compile again. The program size will jump by hundreds of bytes due to the inclusion of the software float library.
        \item \textbf{SRAM vs. Flash:} SRAM is your ``workbench'' - small, fast, and cleared on power-off. Flash is your ``library'' - large, slower, and permanent. We should store all constant, read-only data in the library.
        \item \textbf{PROGMEM Implementation:} The \verb|PROGMEM| keyword is a directive to the compiler to place data in Flash.
\begin{lstlisting}[caption={Code for Module 7.2: Using PROGMEM}]
#include <avr/pgmspace.h> // Required for PROGMEM
#include <stdlib.h>       // For dtostrf
#include <stdio.h>        // For snprintf

// Store this array in Flash memory, not SRAM
const float conversion_factors[] PROGMEM = {
    2.54, 0.3048, 1.60934, 0.453592, 3.78541
};

int main(void) {
    lcd_init(); // Assume LCD driver is included

    // To read from PROGMEM, you MUST use special functions.
    // This copies the float from Flash into a variable in SRAM.
    float feet_to_meters = pgm_read_float(&conversion_factors[1]);

    char lcd_buffer[16];
    char float_buffer[8];
    
    // dtostrf is a non-standard but common AVR function to convert float to string
    // dtostrf(float_val, min_width, num_digits_after_decimal, output_buffer);
    dtostrf(feet_to_meters, 4, 4, float_buffer);
    
    // snprintf is a safe way to format strings
    snprintf(lcd_buffer, 16, ``1ft = %s m'', float_buffer);
    
    lcd_print(lcd_buffer);

    while(1) {}
    return 0;
}
\end{lstlisting}
    \end{enumerate}
    \item \textbf{Checkpoint:} You understand how to use \verb|PROGMEM| to store constant data in Flash and how to use \verb|pgm_read_*| functions to retrieve it, saving precious SRAM.
\end{itemize}
\newpage
\subsection*{Module 7.3: Mathematical Modeling via ODEs (45 Minutes)}

\begin{itemize}
    \item \textbf{Goal:} Reframe complex mathematical functions as initial value problems that can be solved with simple, iterative arithmetic.
    \item \textbf{Conceptual Overview:} We can't afford the memory and processing cost of standard \verb|<math.h>| libraries. We need a way to calculate functions like \verb|sin(x)| or \verb|e^x| ourselves. The answer lies in numerical methods, specifically solving Ordinary Differential Equations (ODEs). The simplest method is Euler's Method.
    \item \textbf{Euler's Method:} An ODE tells us the slope (\verb|y'|) of a function at any point. If we know a starting point \verb|(x0, y0)|, we can approximate the next point by taking a small step in the direction of the slope.
    \begin{center}
    $y_{n+1} = y_n + h \cdot y'(x_n, y_n)$
    \end{center}
    Where \verb|h| is our small step size. The smaller the \verb|h|, the more accurate the result, but the more steps we need.
    \item \textbf{Example: Calculating $e^x$:}
    \begin{itemize}
        \item \textbf{The ODE:} The function $y = e^x$ is defined by the ODE $y' = y$.
        \item \textbf{The Initial Value:} We know that $e^0 = 1$. So, our starting point is $y(0) = 1$.
        \item \textbf{The C Implementation:} Let's approximate $e^{1.0}$ using 100 steps.
    \end{itemize}
\begin{lstlisting}[caption={Code for Module 7.3: Euler's Method for $e^x$}]
float custom_exp(float x) {
    int steps = 100;
    float h = x / steps; // Step size
    float y = 1.0; // Initial value y(0) = 1

    for (int i = 0; i < steps; i++) {
        // For y' = y, the slope is just the current value of y.
        float slope = y;
        // y_new = y_old + h * slope
        y = y + h * slope;
    }
    return y;
}
\end{lstlisting}
    \item \textbf{Checkpoint:} You understand the basic principle of numerical integration: approximating a curve by taking many small, straight-line steps. This is the foundation for building our own math library.
\end{itemize}

\end{document}
